---
title: "Frequency of Steps per Interval"
author: "Daniel Flick"
date: "May 10, 2017"
output:
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Executive Summary

The purpose of this analysis is to analyze the number of steps taken during 5 minute intervals, using an exercise device to track movement.  We first plot the average number of steps (by interval) over a series of users.  We then determine the mean and median steps by interval.  Finally, we impute values for each 'missing' step value in the data set and, based on this updated data set, graphically display the difference in steps taken on the weekdays versus the weekends.

## Step 1 in Analysis 

First, we read in the data set, remove all NA values, and draw a histogram of the frequency of steps taken (during each 5 minute interval).  We then display the mean and median number of steps over all values.

Finally, we graph the mean number of steps taken during each 5 minute interval (across all users) and determine the 5 minute interval associated with the maximum average number of steps.

```{r }

library(lubridate)
library(dplyr)

setwd("C:/Users/djflick/My Documents/R Programming Assignments/Rep_Research")

data<-read.csv("activity.csv",header=T,stringsAsFactors=F)

data$datej<-julian(as.Date(data$date))

data.nona<-data[!is.na(data$steps),]

steps.day<-group_by(data.nona,datej)
data.sumbyday<-summarize(steps.day,total.steps=sum(steps))

hist(data.sumbyday$total.steps, breaks=10, col="blue", xlab="Steps per Day", main="Frequency of Steps per Day")


#Calculate the mean and median number of steps per day
mean.steps<-mean(data.sumbyday$total.steps)
median.steps<-median(data.sumbyday$total.steps)

# Print the mean and median steps per day
print(paste("Mean steps per day is ", mean.steps))
print(paste("Median steps per day is ",median.steps))

#Calculate the number of steps in each interval - and the avg steps per interval - then plot
steps.interval<-group_by(data.nona,interval)
avginterval<-summarize(steps.interval,meansteps.interval=mean(steps))

plot(avginterval$interval,avginterval$meansteps.interval,type='l',xlab="5 Minute Interval", 
     ylab="Mean Steps per Interval",main="Mean Steps per 5 Minute Interval")

#Find the interval with the highest mean number of steps across all days
index.high<-which(avginterval$meansteps.interval==max(avginterval$meansteps.interval))
print(paste("5 Minute Interval associated with max number of steps per day is Interval",avginterval[index.high,1]))

```

## Step 2 in Analysis

In step 2, we first determine the number of observations in the original data set that contain incomplete data.  For each missing value (of the STEPS variable), we impute a step value equal to the average number of steps associated with the associated interval (across all non-missing values).  We then draw a histogram of the frequency of steps, using this updated data set. We note that there is no difference in the mean number of steps using this updated data set (compared to the original).  The MEDIAN of this updated data set is only slightly different than the original data set's MEDIAN number of steps.  

```{r }

print(paste("The number of rows that do not contain complete data is ",sum(!complete.cases(data))))

data.new<-data

for (i in 1:nrow(data))      
 {      ifelse (is.na(data[i,1]), 
        data.new[i,1]<-avginterval[as.integer(which(avginterval$interval==data[i,3])),2],
        data.new[i,1]<-data[i,1])
 }  


#Calculate the number of steps by day, using the data set with imputed values 
steps.day<-group_by(data.new,datej)
data.sumbyday<-summarize(steps.day,total.steps=sum(steps))

#Draw histogram of total steps taken per day

hist(data.sumbyday$total.steps, breaks=10, col="blue", xlab="Steps per Day", main="Frequency of Steps per Day")

#Calculate the mean and median number of steps per day
mean.steps<-mean(data.sumbyday$total.steps)
median.steps<-median(data.sumbyday$total.steps)

# Print the mean and median steps per day
print(paste("Mean steps per day is ", mean.steps))
print(paste("Median steps per day is ",median.steps))

```

## Step 3 in Analysis

Finally, we compare the average number of steps taken across each interval during the weekends and weekdays.  The final chart shows the averages for the two, plotted against each other. 

```{r }


data.new$wkday<-weekdays(as.Date(data.new$date))

data.new$wkday<-ifelse (data.new$wkday=="Saturday" | data.new$wkday=="Sunday","Weekend", "Weekday")

data.new$wkday<-as.factor(data.new$wkday)

datanew.day<-data.new[data.new$wkday=="Weekday",]
datanew.end<-data.new[data.new$wkday=="Weekend",]


steps.dayinterval<-group_by(datanew.day,interval)
steps.endinterval<-group_by(datanew.end,interval)


avgintervalday<-summarize(steps.dayinterval,meansteps.dayinterval=mean(steps))
avgintervalend<-summarize(steps.endinterval,meansteps.endinterval=mean(steps))


plot(avgintervalday$interval,avgintervalday$meansteps.dayinterval,type='l',xlab="5 Minute Interval", 
     ylab="Mean Steps per Interval",main="Mean Steps per 5 Minute Interval",col="red")


lines(avgintervalend$interval,avgintervalend$meansteps.endinterval,type='l',col="blue")
legend("topright",c("Weekday","Weekend"),lwd=c(2.5,2.5),col=c("red","blue"))

```

